Analysis of NLA17 Dissolved Gas Indicator

### 2020-12-11 scripts/dgIndicatorAnalysis.Rmd
Added units to names of some columns, adopted convention of e.gas. for areal emission rate and f.gas for total flux in/out of lake (mass gas/day), added script for spSurvey population estimates (inlcuding extent estimate of N2O undersaturation).

```{r}
# Read data file 
 load(paste0(localPath,
            "/Environmental Protection Agency (EPA)/",
            "ORD NLA17 Dissolved Gas - Documents/",
            "inputData/dg.2020-11-25.RData"))

@@

#  Exploratory spatial analysis
## Population estimates by ecoregion
The survey design allows for population estimates of central tendency and variance by ecoregion (and other subpopulations), which can provide insight into spatial patterning.  There is broad overlap in mean +/- 95% CI estimates for dissolved CH~4~ concentration.  The Northern Plains, Temperate Plains, and Western Mountains have low dissolved CO~2~ concentration.

```{r, fig.cap= "Fig. caption: Mean +/- 95% CI for the dissolved gas concentration indicator"}
# Read population estimates calculated with local neighborhood variance estimate provided by Tom Kincaid (12/12/2019)
pop <- read.csv(file = paste0(localPath, 
                              "/Environmental Protection Agency (EPA)/",
                              "ORD NLA17 Dissolved Gas - Documents/",
                              "inputData/populationEstimates/",
                              "NLA_2017_Percentile_Estimates_Local_Mean_20191212.csv"),
                as.is = TRUE) %>%
  toEPA()  # enforce formatting convention

@@

# look at dissolved concentration indicator
ggplot(pop.eco.diss, aes(estimate, subpopulation)) +
  geom_point() +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct)) +
  facet_wrap(~indicator, scales = "free_x") +
  theme(axis.title = element_blank())

# Read in file containing population estimates for N2O flux (nmol N2O m-2 day-1).
# This file was generated by running Tom Kincaid's code in scripts/CDF_Estimates_jb.Rmd
pop.n2o <- read.csv(file = paste0(localPath, 
                              "/Environmental Protection Agency (EPA)/",
                              "ORD NLA17 Dissolved Gas - Documents/",
                              "inputData/populationEstimates/",
                              "NLA_2017_Percentile_Estimates_Local_Mean_F_N2O_JB20201119.csv"),
                as.is = TRUE) %>%
  toEPA()  # enforce formatting convention

# extract population estimates for ecoregion and nation
pop.n2o <- pop.n2o %>% 
  filter(type %in% c("AG_ECO9_NM", "National"), # extract estimates for ecoregion and nation
         statistic == "Mean") %>% # extract mean estimate
   mutate(subpopulation = factor(subpopulation, 
                             levels = filter(., subpopulation != "National") %>%
                               arrange(estimate) %>%
                               select(subpopulation) %>%
                               pull() %>%
                               c("National", .)))

# look at dissolved concentration indicator
ggplot(pop.n2o, aes(estimate, subpopulation)) +
  geom_point() +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct)) +
  facet_wrap(~indicator, scales = "free_x") +
  theme(axis.title = element_blank())

@@

Population level estimates of variance can be calculated using a simple random sample or local neighborhood estimators.  The local neighborhood estimator will yield lower variance estimates if there is spatial structure in the data.  Below are plotted the 95% confidence intervals calculated using both methods.  In nearly all cases the local neighborhood estimator yields a more restricted interval.

```{r}
# calculate 95% CI using local neighborhood estimator
pop.eco.diss <- pop.eco.diss %>%
  mutate(ci95.local = ucb95pct - lcb95pct)

@@

## Bubble plots
Bubble plots can be used to visualize patterns in geospatial data.  In the plots below, bubble size reflects the gas concentration and color reflects saturation status (i.e. source or sink).  The code immediately below converts the dataframe to a spatial object and reads in the ecoregion ploygons.  The first bubble plot below is of dissolved CO2 which seems to show a pattern of higher concentrations along the eastern seaboard.
```{r}
# To enable spatial analysis of the data, the dataframe will be converted to a 'simple features' (sf) object.
# Define coordinates
coords <- data.frame(longitude = dg$map.lon.dd, latitude = dg$map.lat.dd)

@@

# "Temperate" is misspelled.
ecoR <- ecoR %>% 
  mutate(WSA9_NAME = as.character(WSA9_NAME), # conv to char
         WSA9_NAME = ifelse(WSA9_NAME == "Temporate Plains",
                            "Temperate Plains", # correct sp
                            WSA9_NAME),
         WSA9_NAME = as.factor(WSA9_NAME)) # back to factor

@@

```{r}
# CO2 Bubble plots

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.co2, color = co2.src.snk), 
          show.legend = "point") + # improve legend
  ggtitle("Dissolved CO2 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CO2 (uM)",
             range = c(0.1, 5), # custom size range
             breaks = c(10, 100, 300, 600)) # custom breaks
```

@@

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.ch4, color = ch4.src.snk), 
          show.legend = "point") +
  ggtitle("Dissolved CH4 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CH4 (uM)",
             range = c(0.1, 10), # custom size range
             breaks = c(0.1, 1, 10, 50, 100)) # custom breaks
```

The prevalance of undersaturated dissolved N2O concentrations is an important finding.  The IPCC assumes surface waters are an important source of N2O to the atmosphere.   Webb et al. recently published a paper in PNAS reporting that 70 out of 101 reservoirs sampled in Saskatchewan were N2O sinks.  This paper made quite a splash.  In our study, `r sum(dg$n2o.src.snk == "sink")` out of `r sum(!is.na(dg$dissolved.n2o))` sampled waterbodies were an N2O sink. It is somewhat interesting to note that the some of the highest concentrations were observed in Indiana/Ohio, where I conducted much of my dissertation research on aquatic N2O emissions.

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.n2o, color = n2o.src.snk),
          show.legend = "point") +
  ggtitle("Dissolved N2O (nM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved N2O (nM)",
             range = c(0.1, 10), # custom size range
             breaks = c(1, 10, 25, 50, 100)) # custom breaks
```

We can also look at population estimates of N2O saturation ratio by ecoregion.  Wow, despite the prevalance of N2O undersaturated waterbodies, the 95% CI for the mean N2O saturation ratio encompasses 1 in all ecoregions except the Northern Plains and Northern Appalachians. 
```{r}

# extract saturation indicator, order by n2o
pop.eco.sat <- pop.eco %>%
  filter(grepl("sat", indicator, ignore.case = TRUE)) %>%
  mutate(subpopulation = factor(subpopulation, 
                             levels = filter(pop.eco, indicator == "N2O_SAT_RATIO", subpopulation != "National") %>%
                               arrange(estimate) %>%
                               select(subpopulation) %>%
                               pull() %>%
                               c("National", .)))

# look at dissolved concentration indicator
ggplot(pop.eco.sat, aes(estimate, subpopulation)) +
  geom_point() +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct)) +
  facet_wrap(~indicator, scales = "free_x") +
  theme(axis.title = element_blank()) +
  geom_vline(xintercept = 1)

## Geopackage
Create a simple feature on entire dg data frame, then split out a simple feature, df_sf2, by filtering sample_source = DG.  The geodatabase will contain two simple features, the dissolved gas sampled points/lakes and the nine aggregated ecoregions.
```{r, geopackage, eval=FALSE, echo=FALSE,}
library(janitor) # clean_names function
names(dg)

dg_1 <- select(dg,site.id:map.lon.dd, dissolved.co2, dissolved.ch4, dissolved.n2o, WSA9, WSA9_NAME)
names(dg_1)
dg_1 <- clean_names(dg_1) %>% filter(wsa9 != "AK") # underscores are "safer" names used in GIS
names(dg_1)
# To enable spatial analysis of the data, the dataframe will be converted to a 'simple features' (sf) object.
# Define coordinates
coords <- data.frame(longitude = dg_1$map_lon_dd, latitude = dg_1$map_lat_dd)

dg_sf1 <- st_as_sf(dg_1, coords = c("map_lon_dd", "map_lat_dd"), 
                  crs = 4269) %>% # standard for lat/lon
  st_transform(5070) # 4326 is just the EPSG identifier of WGS84. WGS84 comprises a standard coordinate frame for the Earth, a datum/reference ellipsoid for raw altitude data, and a gravitational equipotential surface (the geoid) that defines the nominal sea level.  5070 is for Albers equal Area.
st_crs(dg_sf1)

dg_sf1 <- filter(dg_sf1, sample_source == "DG") # now 1219 obs vs. previously 1185 obs., why change in sample size
st_crs(dg_sf1)

@@

# write out dg.sf and ecoR as two layers in geopackage, gpkg, geodatabase
# 74. Creating a geopackage geodatabase, then adding more layers
# https://github.com/r-spatial/sf/issues/392
# creating a _v1.gpkg having large sample size and should be in Albers
st_write(dg_sf1, dsn = file.path(getwd(), "nladg_v1.gpkg"), layer = "dg_sf1", driver = "GPKG", quiet = FALSE)
st_layers("nladg_v1.gpkg")

# https://gis.stackexchange.com/questions/223240/writing-multiple-layers-to-geopackage-using-writeogr-in-r
st_write(ecoR1, dsn = file.path(getwd(), "nladg_v1.gpkg"), layer = "ecoR1", driver = "GPKG", append = TRUE, quiet = FALSE)
st_layers("nladg_v1.gpkg")
#CPL polygon not appearing in geopackage

names(dg_sf1)
summary(dg_sf1$dissolved_n2o) # matches Roy's stats
```

@@

# Unsure that Springsteen function below will work as euc is an spDists created object that I think retained ecoregional & not sure that's so with rdistmat1
# Add within-region summary statistics to summary table 
for (r in 1:length(ecoRname1)){ # loop through regions and fill in the matrices
  euc.reg <- euc[grep(ecoRname1[r], rownames(euc)), grep(ecoRname1[r],colnames(euc))] # retrieving only columns and rows in the distance matrix that contain the same region code in their column/row name
  # retrieving only columns and rows in the distance matrix that contain the same region code in their column/row name
  euc.summary[, (r+1)] <- round(c( min(euc.reg[euc.reg!=0]), quantile(euc.reg[euc.reg!=0], probs=seq(0,1, 0.25), names=F)[2], median(euc.reg[euc.reg!=0]), quantile(euc.reg[euc.reg!=0], probs=seq(0,1, 0.25), names=F)[4], max(euc.reg[euc.reg!=0]) ),1)
}

@@

# GitHub question:  if I save a workspace, .RData, will that be pushed when I push to GitHub?  Or, should data frames go in .gitignore?
```

## Lagged scatterplots
Note that can use simple feature object, such as cpl, for histogram, but spatial object needed for lagged scatterplots.  I took log10 transform to  get data somewhat unimodal and symmetric.
```{r, eval=FALSE, echo=FALSE, laggedscatterplot}
# cpl
cpl_sp <- as(cpl, "Spatial")
class(cpl_sp)
ggplot(cpl,aes(dissolved.ch4)) + geom_histogram()
ggplot(cpl,aes(dissolved.ch4)) + geom_histogram() + scale_x_log10()
hscat(log10(dissolved.ch4)~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(dissolved.ch4~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(dissolved.ch4~1, cpl_sp, c(0, 10000, 25000, 50000, 150000, 250000, 500000))

ggplot(cpl,aes(dissolved.co2)) + geom_histogram()
ggplot(cpl,aes(dissolved.co2)) + geom_histogram() + scale_x_log10()

hscat(log10(dissolved.co2)~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(dissolved.co2~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(log10(dissolved.co2)~1, cpl_sp, c(0, 10000, 25000, 50000, 150000, 250000, 500000))

ggplot(cpl,aes(dissolved.n2o)) + geom_histogram()
ggplot(cpl,aes(dissolved.n2o)) + geom_histogram() + scale_x_log10()

hscat(log10(dissolved.n2o)~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(dissolved.n2o~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

>>>>

```{r}
# Read data file 
 load(paste0(localPath,
            "/Environmental Protection Agency (EPA)/",
            "ORD NLA17 Dissolved Gas - Documents/",
            "inputData/dg.2020-12-02.RData"))

@@

#  Exploratory spatial analysis
## Population estimates by ecoregion
### Dissolved gas concentration
The survey design allows for population estimates of central tendency and variance by ecoregion (and other subpopulations), which can provide insight into spatial patterning.  There is broad overlap in mean +/- 95% CI estimates for dissolved CH~4~ and N~2~O concentration across ecoregions.  The Northern Plains, Temperate Plains, and Western Mountains have low dissolved CO~2~ concentration.

```{r, fig.cap= "Fig. caption: Mean +/- 95% CI for the dissolved gas concentration indicator"}
# Read population estimates calculated with local neighborhood variance estimate.  Code provided by Tom Kincaid 
# (12/12/2019), then modified to accomdate the gas emission rate and flux variables.  See CDF_Estimates_Jb.Rmd.
pop <- read.csv(file = paste0(localPath, 
                              "/Environmental Protection Agency (EPA)/",
                              "ORD NLA17 Dissolved Gas - Documents/",
                              "inputData/populationEstimates/",
                              "NLA_2017_Percentile_Estimates_Local_Mean_JB20201202.csv"),
                as.is = TRUE) %>%
  toEPA()  # enforce formatting convention

@@

# look at dissolved concentration indicator
ggplot(pop.eco.diss, aes(estimate, subpopulation)) +
  geom_point() +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct)) +
  facet_wrap(~indicator, scales = "free_x") +
  theme(axis.title = element_blank())
```

### Dissolved gas saturation ratios and source/sink status
Dissolved gas saturation ratios are calculated as the ratio of the observed to equilibrium concentration.  The equilibrium concentration (sometimes called the saturation concentration) is the concentration expected if the waterbody were in equilibrium with the atmosphere.  It is calculated from the partial pressure of the gas in the atmosphere, water temperature, and the gas specific solubility.  Saturation ratios >1 indicate the water body is supersaturated and is a source of the gas to the atmosphere.  Saturation ratios <1 indicate that the waterbody is undersaturated and is consuming the gas from the atmosphere.  The saturation ratio can therefore be used to categorize each waterbody as a 'sink' or a 'source'.

``` {r}
# read in source/sink population data
# Read extent estimates generated with CDF_Estimates_Jb.Rmd.  File provided by Tom Kincaid (12/12/2019), then modified to accomodate the gas emission rate, flux variables, and cat.analysis for N2O source/sink.  
n2o.extent <- read.csv(file = paste0(localPath, 
                                     "/Environmental Protection Agency (EPA)/",
                                     "ORD NLA17 Dissolved Gas - Documents/",
                                     "inputData/populationEstimates/",
                                     "NLA_2017_Extent_Estimates20201203.csv"),
                       as.is = TRUE) %>%
  toEPA() %>% # enforce formatting convention
  filter(type %in% c("AG_ECO9_NM", "National"), # extract estimates for ecoregion and nation
         category == "sink") %>% # extract extent estimate for sink
  mutate(subpopulation = factor(subpopulation, # arrange by extent estimate for sink
                                levels = filter(., subpopulation != "National") %>%
                                  arrange(estimate.p) %>%
                                  select(subpopulation) %>%
                                  pull() %>%
                                  c("National", .)))
```

The binary 'source' or 'sink' values can be analyzed at the population level using the cat.analysis function in the spsurvey package. The function provides an 'extent' estimate, which is the proportion of waterbodies in each region that are functioning as a source or sink. Over half of the waterbodies are functioning as N2O sinks in all ecoregions.  At the national scale, `r n2o.extent %>% filter(type == "National") %>% select(estimate.p) %>% pull() %>% round(., 1)`% (95% CI: `r n2o.extent %>% filter(type == "National") %>% select(lcb95pct.p) %>% pull() %>% round(., 1)` - `r n2o.extent %>% filter(type == "National") %>% select(ucb95pct.p) %>% pull() %>% round(., 1)`) of lakes and reservoirs were net N~2~O sinks during the index period, suggesting that US lakes and reservoirs are a net N~2~O sink.  However, the 95% confidence intervals of the mean N~2~O saturation ratio include 1 for all ecoregions except the Northern plains which are undersaturated.  At the national scale, the mean N~2~O saturation ratio indicates supersaturation (`r pop.eco %>% filter(type == "National", statistic == "Mean", indicator == "N2O_SAT_RATIO") %>% select(estimate) %>% pull() %>% round(., 2)`% (95% CI: `r pop.eco %>% filter(type == "National", statistic == "Mean", indicator == "N2O_SAT_RATIO") %>% select(lcb95pct) %>% pull() %>% round(., 2)` - `r pop.eco %>% filter(type == "National", statistic == "Mean", indicator == "N2O_SAT_RATIO") %>% select(ucb95pct) %>% pull() %>% round(., 2)`)), suggesting that US lakes and reservoirs function as a net N~2~O sink.  These findings indicate that high levels of N~2~O saturation in a few waterbodies offset undersaturation in most waterbodies, yielding a mean saturation ratio > 1.

```{r message=FALSE}

## Bubble plot of observed N2O saturation index----------------
# To enable spatial analysis of the data, the dataframe will be converted to a 'simple features' (sf) object.
# Define coordinates
coords <- data.frame(longitude = dg$map.lon.dd, latitude = dg$map.lat.dd)

# Check CRS
#st_crs(ecoR) # 3857
ecoR <- st_transform(ecoR, 5070) # convert to CONUS Albers
#st_crs(ecoR) # 5070

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = n2o.sat.ratio, color = n2o.src.snk),
          show.legend = "point") +
  ggtitle("N2O saturation ratio") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size_binned(name = "N2O saturation ratio",
             range = c(1, 8), # custom size range
             breaks = c(0.1, 0.5, 1, 2))  # custom breaks

## horizontal bar chart of N2O sink extent---------------
ggplot(n2o.extent, aes(estimate.p, subpopulation)) +
  geom_bar(stat = "identity", aes(fill = subpopulation, color = subpopulation), alpha = 0.25) +
  geom_errorbarh(aes(xmax = ucb95pct.p, xmin = lcb95pct.p, color = subpopulation), height = 0, size = 1) +
  xlab("Percent of waterbodies functioning as N2O sinks") +
  theme(axis.title.y = element_blank())

## horizontal bar chart of N2O saturation ratio-------------
# extract saturation indicator, order by n2o extent (see above).  This enforces same order for both plots.
pop.eco.sat <- pop.eco %>%
  filter(grepl("sat", indicator, ignore.case = TRUE)) %>%
  mutate(subpopulation = factor(subpopulation, 
                             levels = filter(n2o.extent, subpopulation != "National") %>%
                                  arrange(estimate.p) %>%
                                  select(subpopulation) %>%
                                  pull() %>%
                               as.character() %>%
                                  c("National", .)))

# N2O saturation ratio; horizontal bar chart
ggplot(filter(pop.eco.sat, indicator == "N2O_SAT_RATIO"), aes(estimate, subpopulation)) +
  geom_bar(stat = "identity", aes(fill = subpopulation, color = subpopulation), alpha = 0.25) +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct, color = subpopulation), height = 0, size = 1) +
  xlab("N2O saturation ratio") +
  theme(axis.title.y = element_blank()) +
  geom_vline(xintercept = 1)

@@

### Emission rates and flux
The emission rate (E~gas~) is the rate at which a gas crosses the air water interface and is expressed on an areal basis (i.e. mass m^-2^ day^-1^).  It is calculated as a function of the dissolved gas concentration and gas transfer velocity (see dataMunge.Rmd).  The 95% confidence interval of the mean N~2~O emission rate is quite large in several ecoregions, encompassing 0 in all but the Northen Plains and Xeric ecoregions.  The mean N~2~O emission rate at the national scale is positive (`r  pop.eco %>% filter(indicator == "E_N2O_NMOL_D", subpopulation == "National", statistic == "Mean") %>% select(estimate) %>% pull() %>% round(., 1)/1000` umol N~2~O m^-2^ day^-1^), although the 95% confidence interval includes 0, similar to the N~2~O saturation ratio pattern.

```{r, fig.cap= "Fig. caption: Mean +/- 95% CI for the emission rate (mass of GHG time-1 area-1)"}
# extract population estimates for ecoregion and nation
pop.eco.emission <- pop.eco %>% 
  filter(grepl("^E", indicator)) %>% # ^ means starts with.  Extract the emission rates 
  # order by N2O
   mutate(subpopulation = factor(subpopulation, 
                             levels = filter(., indicator == "E_N2O_NMOL_D", subpopulation != "National") %>%
                               arrange(estimate) %>%
                               select(subpopulation) %>%
                               pull() %>%
                               c("National", .)))

# look at dissolved concentration indicator
ggplot(filter(pop.eco.emission, indicator == "E_N2O_NMOL_D"), 
       aes(estimate/1000, subpopulation)) + # convert to umol
  geom_point(aes(color = subpopulation, fill = subpopulation), alpha = 0.25, size = 5) +
  geom_errorbarh(aes(xmax = ucb95pct/1000, xmin = lcb95pct/1000, color = subpopulation), height = 0, size = 1) + # convert to umol
  geom_vline(xintercept = 0) +
  xlab("N2O emission rate (umol N2O m-2 day-1") + # converted from nmol to umol (see above)
  theme(axis.title.y = element_blank())

The N~2~O flux at a waterbody is the total mass of N~2~O crossing the air-water interface per unit time (e.g. mass N~2~O day^-1^) and is calculated as the product of the emission rate (mass N~2~O m^-2^ day^-1^) and waterbody area (m^2^).  Negative values indicate N2O invasion into the waterbody and positive values indicate N~2~O efflux from the waterbody to the atmosphere.  An assessment of the role of lakes and reservoirs in the atmospheric N~2~O budget must be based on flux, rather than emission rates, because flux accounts for waterbody size.  For example, N~2~O emissions from several small waterbodies could be completely offset by N~2~O uptake by a single large waterbody with a small negative emission rate (e.g. N~2~O sink), yet the mean emission rate for these waterbodies would be > 0.

The mean N~2~O flux per waterbody is close to zero for most ecoregions and the 95% confidence interval includes zero for all ecoregions.  The national scale mean N~2~O flux is slightly negative (`r  pop.eco %>% filter(indicator == "F_N2O_M_D", subpopulation == "National", statistic == "Mean") %>% select(estimate) %>% pull() %>% round(., 2)` mol N~2~O day^-1^), but the 95% CI includes 0.

```{r, fig.cap= "Fig. caption: Mean +/- 95% CI of the GHG flux across the air-water interface (mass GHG / time)"}
# extract population estimates for ecoregion and nation
pop.eco.flux <- pop.eco %>% 
  filter(grepl("^F", indicator)) %>% # ^ means starts with.  Extract the flux rates 
  # order by N2O
   mutate(subpopulation = factor(subpopulation, 
                             levels = filter(., indicator == "F_N2O_M_D", subpopulation != "National") %>%
                               arrange(estimate) %>%
                               select(subpopulation) %>%
                               pull() %>%
                               c("National", .)))

# look at dissolved concentration indicator
ggplot(filter(pop.eco.flux, indicator == "F_N2O_M_D"),
       aes(estimate, subpopulation)) +
  geom_point(aes(color = subpopulation, fill = subpopulation), alpha = 0.25, size = 5) +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct, color = subpopulation), height = 0, size = 1) +
  geom_vline(xintercept = 0) +
  xlab("N2O flux (mol N2O day^-1 waterbody^-1") +
  theme(axis.title.y = element_blank())

### 13C paper
```{r}
# 13C; horizontal bar chart by ecoregion
pop.eco.13c <- filter(pop.eco, indicator == "DELTA_13_CH4_C")
ggplot(pop.eco.13c, aes(estimate, subpopulation)) +
  geom_bar(stat = "identity", aes(fill = subpopulation, color = subpopulation), alpha = 0.25) +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct, color = subpopulation), height = 0, size = 1) +
  theme(axis.title.y = element_blank())

# 13C; horizontal bar chart by origin
pop.origin.13c <- filter(pop, type == "LAKE_ORGN", statistic == "Mean", indicator == "DELTA_13_CH4_C")
ggplot(pop.origin.13c, aes(estimate, subpopulation)) +
  geom_bar(stat = "identity", aes(fill = subpopulation, color = subpopulation), alpha = 0.25) +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct, color = subpopulation), height = 0, size = 1) +
  theme(axis.title.y = element_blank())

# 13C; horizontal bar by size
pop.size.ch4 <- filter(pop, type == "AREA_CAT6", statistic == "Mean", indicator == "DISSOLVED_CH4")
ggplot(pop.size.ch4, aes(estimate, subpopulation)) +
  geom_bar(stat = "identity", aes(fill = subpopulation, color = subpopulation), alpha = 0.25) +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct, color = subpopulation), height = 0, size = 1) +
  theme(axis.title.y = element_blank())

# 13CH4 Bubble plots

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = delta.13.ch4), 
          show.legend = "point") + # improve legend
  ggtitle("13CH4") +
  scale_size(name = "13C-CH4",
             range = c(0.05, 3), # custom size range
             breaks = c(-80, -50, -25, 0)) # custom breaks

Similarly, we can look at population level estimates of GHG emission rates (mass of GHG time^-1^ area^-1^).

We can also look at the GHG flux across the air-water interface (mass GHG time^-1^)

## Simple random sample vs local neighborhood variance estimator
Population level estimates of variance can be calculated using a simple random sample or local neighborhood estimators.  The local neighborhood estimator will yield lower variance estimates if there is spatial structure in the data.  Below are plotted the 95% confidence intervals calculated using both methods.  In nearly all cases the local neighborhood estimator yields a more restricted interval.

```{r, message=FALSE}
# calculate 95% CI using local neighborhood estimator
pop.eco.diss <- pop.eco.diss %>%
  mutate(ci95.local = ucb95pct - lcb95pct)

@@

## Bubble plots
Bubble plots can be used to visualize patterns in geospatial data.  In the plots below, bubble size reflects the gas concentration and color reflects saturation status (i.e. source or sink).  The code immediately below converts the dataframe to a spatial object and reads in the ecoregion ploygons.  The first bubble plot below is of dissolved CO2 which seems to show a pattern of higher concentrations along the eastern seaboard.
```{r, message=FALSE}
# To enable spatial analysis of the data, the dataframe will be converted to a 'simple features' (sf) object.
# Define coordinates
coords <- data.frame(longitude = dg$map.lon.dd, latitude = dg$map.lat.dd)

@@

@@

```{r, message=FALSE}
# CO2 Bubble plots

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.co2.umol, color = co2.src.snk), 
          show.legend = "point") + # improve legend
  ggtitle("Dissolved CO2 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CO2 (uM)",
             range = c(0.1, 5), # custom size range
             breaks = c(10, 100, 300, 600)) # custom breaks
```

@@

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.ch4.umol, color = ch4.src.snk), 
          show.legend = "point") +
  ggtitle("Dissolved CH4 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CH4 (uM)",
             range = c(0.1, 10), # custom size range
             breaks = c(0.1, 1, 10, 50, 100)) # custom breaks
```

The prevalance of undersaturated dissolved N2O concentrations is an important finding.  The IPCC assumes surface waters are an important source of N2O to the atmosphere.   Webb et al. recently published a paper in PNAS reporting that 70 out of 101 reservoirs sampled in Saskatchewan were N2O sinks.  This paper made quite a splash.  In our study, `r sum(dg$n2o.src.snk == "sink")` out of `r sum(!is.na(dg$dissolved.n2o.nmol))` sampled waterbodies were an N2O sink. It is somewhat interesting to note that the some of the highest concentrations were observed in Indiana/Ohio, where I conducted much of my dissertation research on aquatic N2O emissions.

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.n2o.nmol, color = n2o.src.snk),
          show.legend = "point") +
  ggtitle("Dissolved N2O (nM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved N2O (nM)",
             range = c(0.1, 10), # custom size range
             breaks = c(1, 10, 25, 50, 100)) # custom breaks
```

## Geopackage
Create a simple feature on entire dg data frame.  The geopackage will contain two simple features, the dissolved gas sampled points/lakes and the nine aggregated ecoregions.
```{r, geopackage, eval=FALSE, echo=FALSE,}
library(janitor) # clean_names function
names(dg)

dg_1 <- select(dg,site.id:map.lon.dd, dissolved.co2.umol, dissolved.ch4.umol, dissolved.n2o.nmol, WSA9, WSA9_NAME)
names(dg_1)
dg_1 <- clean_names(dg_1) %>% filter(wsa9 != "AK") # underscores are "safer" names used in GIS
names(dg_1)
# To enable spatial analysis of the data, the dataframe will be converted to a 'simple features' (sf) object.
# Define coordinates
coords <- data.frame(longitude = dg_1$map_lon_dd, latitude = dg_1$map_lat_dd)

dg_sf1 <- st_as_sf(dg_1, coords = c("map_lon_dd", "map_lat_dd"), 
                  crs = 4269) %>% # standard for lat/lon
  st_transform(5070) # 4326 is just the EPSG identifier of WGS84. WGS84 comprises a standard coordinate frame for the Earth, a datum/reference ellipsoid for raw altitude data, and a gravitational equipotential surface (the geoid) that defines the nominal sea level.  5070 is for Albers equal Area.
st_crs(dg_sf1)
dim(dg_sf1) # 1184 observations

@@

# write out dg.sf and ecoR as two layers in geopackage, gpkg, geodatabase
# 74. Creating a geopackage geodatabase, then adding more layers
# https://github.com/r-spatial/sf/issues/392
# creating a _v1.gpkg having large sample size and should be in Albers
st_write(dg_sf1, dsn = file.path(getwd(), "nladg_v1.gpkg"), layer = "dg_sf1", driver = "GPKG", quiet = FALSE, 
         delete_layer = TRUE) # needed, else function gives error if object already exists
st_layers("nladg_v1.gpkg")

# https://gis.stackexchange.com/questions/223240/writing-multiple-layers-to-geopackage-using-writeogr-in-r
st_write(ecoR1, dsn = file.path(getwd(), "nladg_v1.gpkg"), layer = "ecoR1", driver = "GPKG", append = TRUE, quiet = FALSE, delete_layer = TRUE) # needed, else function gives error if object already exists
st_layers("nladg_v1.gpkg")

names(dg_sf1)
summary(dg_sf1$dissolved_n2o_nmol) # matches Roy's stats
```

@@

# Unsure that Springsteen function below will work as euc is an spDists created object that I think retained ecoregional & not sure that's so with rdistmat1
# Add within-region summary statistics to summary table 
# for (r in 1:length(ecoRname1)){ # loop through regions and fill in the matrices
#   euc.reg <- euc[grep(ecoRname1[r], rownames(euc)), grep(ecoRname1[r],colnames(euc))] # retrieving only columns and rows in the distance matrix that contain the same region code in their column/row name
#   # retrieving only columns and rows in the distance matrix that contain the same region code in their column/row name
#   euc.summary[, (r+1)] <- round(c( min(euc.reg[euc.reg!=0]), quantile(euc.reg[euc.reg!=0], probs=seq(0,1, 0.25), names=F)[2], median(euc.reg[euc.reg!=0]), quantile(euc.reg[euc.reg!=0], probs=seq(0,1, 0.25), names=F)[4], max(euc.reg[euc.reg!=0]) ),1)
# }

@@

# GitHub question:  if I save a workspace, .RData, will that be pushed when I push to GitHub?  Or, should data frames go in #.gitignore?
# Jake - we are using a public github repo under the EPA enterprise license.  Anything we push to github can be downloaded by the public.  We should use the shared documents library at our SharePoint site for data.  Code can go to github.
```

## Lagged scatterplots
Note that can use simple feature object, such as cpl, for histogram, but spatial object needed for lagged scatterplots.  I took log10 transform to  get data somewhat unimodal and symmetric.
```{r, eval=FALSE, echo=FALSE, laggedscatterplot}
# cpl
cpl_sp <- as(cpl, "Spatial")
class(cpl_sp)
ggplot(cpl,aes(dissolved.ch4.umol)) + geom_histogram()
ggplot(cpl,aes(dissolved.ch4.umol)) + geom_histogram() + scale_x_log10()
hscat(log10(dissolved.ch4.umol)~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(dissolved.ch4.umol~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(dissolved.ch4.umol~1, cpl_sp, c(0, 10000, 25000, 50000, 150000, 250000, 500000))

ggplot(cpl,aes(dissolved.co2.umol)) + geom_histogram()
ggplot(cpl,aes(dissolved.co2.umol)) + geom_histogram() + scale_x_log10()

hscat(log10(dissolved.co2.umol)~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(dissolved.co2.umol~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))

hscat(log10(dissolved.co2.umol)~1, cpl_sp, c(0, 10000, 25000, 50000, 150000, 250000, 500000))

ggplot(cpl,aes(dissolved.n2o.nmol)) + geom_histogram()
ggplot(cpl,aes(dissolved.n2o.nmol)) + geom_histogram() + scale_x_log10()

hscat(log10(dissolved.n2o.nmol)~1, cpl_sp, c(0, 50000, 150000, 250000, 500000, 1000000, 1500000))
